---
title: "May Event - Demo - htmlwidgets for R"
author: "R-Ladies Coding Club (London)"
date: "17 May 2016"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

> *This is the script for the htmlwidgets for R demo at the R-Ladies Coding Club's May Event. Published doc available on RPubs: http://rpubs.com/crt34/widgetdemo*  

---

#### Demo of selected htmlwidgets
-- A. Time-Series Plot     
-- B. Heatmap    
-- C. 3D Scatterplot      
-- D. 3D Globe    

#### What are htmlwidgets for R?
http://www.htmlwidgets.org  

A growing number of htmlwidget R libraries have been developed from popular JavaScript charting libraries, such as D3.js, plotly.js, Leaflet.js etc.  

This htmlwidgets framework/interface therefore allows R users to generate interactive JavaScript data visualisations using R code, through the normal operation of loading relevant libraries and calling appropriate functions.

These interactive plots can be displayed in RStudio, embedded in R markdown documents and Shiny web apps, and shared in standalone html files over email, dropbox etc.

#### What Use-Cases do we have for htmlwidgets for R?
-- Data Storytelling: presenting & communicating findings   
-- Exploratory Data Analysis (EDA) 

---

#### Install Packages for all Demos

```{r install all packages, eval=FALSE}
install.packages("dygraphs")
install.packages("xts")
install.packages("d3heatmaps")
install.packages("threejs")
```

### A. Time-Series Plot
Produce an interactive time-series visualisation using the `dygraphs` R library. Objects for plotting must be in `xts` format (extensible time series).

Time-series data = Australian wine sales in no. of bottles, by wine makers, between Jan 1980 - Aug 1994.   

Using the `stl` function to perform the seasonal trend decomposition, which extracts the seasonal component, trend, component and remainder from the original data.

#### (i) Call dygraph & xts libraries

```{r install and call dygraph, message=FALSE}
library(dygraphs)
library(xts)
```

#### (ii) Load Data
In this instance the Australian Wine dataset `wineind` being used is from the `forecast` package so we load this first, then call the data.    

(Original source: http://data.is/TSDLdemo)

```{r load forecast and data, message=FALSE}
library(forecast)
data("wineind")
```

#### (iii) Explore Data
```{r explore wineind, eval = FALSE}
?wineind
class(wineind)
```

```{r view wineind, eval = FALSE}
wineind
```

```{r plot wineind}
plot(wineind) #base R plot
```

#### (iv) Apply Decomposition algorithm

Apply `stl` algorithm, which divides the time series into 3 components, trend, seasonality & remainder, using Loess (a method for estimating nonlinear relationships).

`stl` documentation: https://stat.ethz.ch/R-manual/R-devel/library/stats/html/stl.html 

```{r stl_wine}
stl_wine <- stl(wineind, s.window = "periodic") #s.window controls variation of seasonal component
```

```{r stl_wine data, eval=FALSE}
stl_wine
```

```{r stl_wine plot}
plot(stl_wine) #base R plot
```

#### (v) Prepare decomposed data as a single xts object
First, convert each decomposed component into a separate `ts` (time series) object, and then into `xts` (extensible time series) object:

```{r convert to xts}
#(1980, 1) is start year & unit of observations, 12 is the no. of observations per period i.e. monthly

random_stl <- ts(stl_wine$time.series[,3], start = c(1980, 1), frequency=12)
random_stl.xts <- as.xts(random_stl)

seasonal_stl <- ts(stl_wine$time.series[,1], start = c(1980, 1), frequency=12)
seasonal_stl.xts <- as.xts(seasonal_stl)

trend_stl <- ts(stl_wine$time.series[,2], start = c(1980, 1), frequency=12)
trend_stl.xts <- as.xts(trend_stl)
```

Secondly, column bind the 3 `xts` components together to create a single `xts` object for plotting:

```{r 3 to 1}
wine.plot <- cbind(random_stl.xts, seasonal_stl.xts, trend_stl.xts)

#re-name columns
colnames(wine.plot) <- c("remainder", "seasonal", "trend")
```

#### (vi) Create Dygraph
Basic:
```{r uncustomised dygraph, eval=FALSE}
dygraph(wine.plot) %>%
    dyOptions(stackedGraph = TRUE)
```

#### (vii) Customise Dygraph
Add Title & Range Selector & change colours:
```{r range, eval=FALSE}
dygraph(wine.plot, main = "Australian Wine Sales") %>%
    dyOptions(stackedGraph = TRUE, colors=RColorBrewer::brewer.pal(3, "Set1"))%>%
      dyRangeSelector()
```

---

### B. Heatmap
Produce a D3 heatmap using the `d3heatmap` R library.

#### (i) Call d3heatmap library

```{r install and call heatmaps}
library(d3heatmap)
```

#### (ii) Load Data
Data created from personal burger blog, in `wide` format necessary to use this library. (https://burgerite.blogspot.co.uk)    

Data available here: https://github.com/RLadiesCodingLondon/MayEvent-Chiin-htmlwidgets

```{r load burger data, eval=FALSE}
burger <- read.csv("burger.csv", header = TRUE, row.names = 1)
```

#### (iii) Explore Data
5 attributes, 25 burgers reviewed.
```{r explore burger, eval=FALSE}
View(burger)
```

#### (iv) Create Heatmap
Blue = higher score, Red = lower score
```{r uncustomised heatmap, eval=FALSE}
d3heatmap(burger)
```

---

### C. 3D Scatterplot
Produce a 3D scatterplot using the `threejs` R library using same `burger` data.

#### (i) Call threejs library

```{r install and call scatterplot}
library(threejs)
```

#### (ii) Create Scatterplot

```{r thing scatterplot, eval=FALSE}
scatterplot3js(burger$proportions, burger$ingredients, burger$patty, flip.y=FALSE)
#flip.y to control direction of y-axis
```

#### (iii) Customise Scatterplot

```{r other thing scatterplot, eval=FALSE}
label <- c("proportions", "patty", "ingredients") #define axis labels

scatterplot3js(burger$proportions, burger$ingredients, burger$patty, 
               axisLabels = label,
               labels=row.names(burger),
               flip.y = FALSE,
               color=rainbow(length(burger$value)),
               stroke = NULL,
               size=burger$value/5,
               renderer = "canvas")
```

---

### D. 3D Globe
Produce a 3D globe, again using the `threejs` R library.

#### (i) Load Data
Geo-location data from Google Analytics of burger blog traffic.

Data available here: https://github.com/RLadiesCodingLondon/MayEvent-Chiin-htmlwidgets
```{r load globe data}
views <- read.csv("globejs.csv", header = TRUE, row.names = 1)
```

#### (ii) Explore Data
169 cities with cumulative blog page views >= 2 since Sep '15.
```{r explore traffic, eval=FALSE}
View(views)
```

#### (iii) Create Globe

```{r unustomised globe, eval=FALSE}
globejs(lat = views$Lat, long = views$Long, value = views$PageViews)
```

#### (iv) Customise Globe

```{r customised globe, eval=FALSE}
traffic <- 10000*views$PageViews/max(views$PageViews)

globejs(lat = views$Lat, long = views$Long, value = traffic,
        atmosphere = TRUE,
        pointsize = 0.5)
```
