---
title: An introduction to R
subtitle: Beginners drop-in session
author: R-ladies London team | @RladiesLondon
date: 26th October 2016
output: 
  ioslides_presentation:
    smaller: true
    css: custom.css
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, 
                      eval = FALSE)
```

# Basics

## Dependencies

You MUST have the following installed to participate:

> - R [https://www.r-project.org/](https://www.r-project.org/)

> - RStudio [https://www.rstudio.com/](https://www.rstudio.com/) <br><br> product -> Rstudio -> Desktop -> Open Source Edition<br>

# What's the plan for today?

## Agenda

* What's R?

* What's RStudio? Why do we need it? Is R not enough?

* R documentation pages, Rseek and Stackoverflow

* Load & explore a dataset

* Basic plots

# Let's start!

## What's R?

## What's RStudio? Why do we need it? Is R not enough?

## R documentation pages, Rseek and Stackoverflow
How to get to the in-built help pages
```{r} ?boxplot 
```

## Setting your working directory
Check where your working directory is- this is the location where all files and functions will be read and written to
```{r}
getwd()
```
If this isn't where your files are then you can set it through (put your own path in the quotes!)
```{r eval=FALSE} setwd("C:\Documents\mypath") 
```
Or in Rstudio you can go: Session>Set Working Directory>Choose Directory and navigate to your folder

## Load & explore a dataset

Have a look at the in-built data sets in R
```{r eval=FALSE}
data()
```

#Look at the in-built data set on trees
```{r eval=FALSE}
trees
```

#Explore the trees data set
The top of the data ```{r} head(trees)```
The end of the data set ```{r} tail(trees) ```
The size and type of the data ```{r} str(trees) ```
Summary statistics on each of the fields ```{r} summary(trees) ```
Pull out only one of the fields ```{r} summary(trees$Girth) ```

## Basic plots

Create your first plot
```{r}
boxplot(trees)
```
Create another variable with only one of the fields and plot that
```{r}
treeGirth<-trees$Girth
boxplot(treeGirth)
boxplot(treeGirth,xlab="Girth (units)",title="Tree Girths")
```

## Where to go next
It would be best if you can get your own data and start exploring that.
Make sure that your working directory is set to your file's location
```{r eval=FALSE} data<-read.csv("myfile.csv")
```

Great tutorials:
edx MiT course: https://www.edx.org/course/analytics-edge-mitx-15-071x-2
DataCamp: https://www.datacamp.com/
Coursera: https://www.coursera.org/learn/r-programming
Great Kaggle Tutorials: https://www.kaggle.com/mrisdal/titanic/exploring-survival-on-the-titanic

